### 2022-02-23 ehoh/ehoh.tex
inference rules
finished

A side effect of the flattened representation is that prefix subterms are not
shared. For example, the terms $\cst{f}\;\cst{a}$ and $\cst{f}\;\cst{a}\;\cst{b}$
correspond to the flattened cells $\cst{f}(\cst{a})$ and
$\cst{f}(\cst{a}, \cst{b}).$ The argument subterm $\cst{a}$ is shared, but not
the prefix $\cst{f}\;\cst{a}.$ Similarly,
$x$ and $x\;\cst{b}$ are represented by two distinct cells, $x()$ and
$x(\cst{b})$, and there is no connection between the two occurrences of~$x$.
%
In particular,
despite perfect sharing, their \verb|binding| fields are unconnected, leading
to inconsistencies.

@@

\noindent
\begin{tabular}{ll}
  \textsf{Delete} & $\{ t \unif t \} \uplus \mathit{S} \Unifarrow \mathit{S}$ \\[\jot]
  \textsf{Decompose} & $\{ \cst{f} \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{m} \} \uplus \mathit{S} \Unifarrow {} 
                        \mathit{S} \cup \{ s_1 \unif t_1, \ldots, s_m \unif t_m \}$ \\[\jot]
  \textsf{DecomposeX} & $\{ x \; \tuple{s}{{m}} \unif u \; \tuple{t}{{m}} \} \uplus \mathit{S} \Unifarrow {} 
                         \mathit{S} \cup \{ x \unif u{,}\; s_1 \unif t_1, \ldots, s_{m} \unif t_{m} \}$  \\
                      & if $x$ and $u$ have the same type and $m > 0$ \\[\jot]
  \textsf{Orient}     & $\{ \cst{f} \; \overline{s} \unif x \; \overline{t} \} \allowbreak \uplus \mathit{S}
                         \Unifarrow \mathit{S} \cup \{ x \; \overline{t} \unif \cst{f} \; \overline{s} \}$ \\[\jot]
  \textsf{OrientXY}   &  $\{ x \; \tuple{s}{m} \unif y \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S}
                         \Unifarrow \mathit{S} \cup \{ y \; \tuple{t}{n} \unif x \; \tuple{s}{m} \}$  \\
                      &   if $m > n$ \\[\jot]
  \textsf{Eliminate}  & $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \{ x \unif t \} \cup \{ x \mapsto t \}(\mathit{S})$ \\
                      &   if $x \in \Var(\mathit{S}) \setminus \Var(t)$
\end{tabular}

@@

\noindent
\begin{tabular}{ll}
  \textsf{Clash} &   $\{ \cst{f} \; \overline{s} \unif \cst{g} \; \overline{t} \} \uplus \mathit{S} \Unifarrow \bot$; if $\cst{f} \not= \cst{g}$ \\[\jot]
  \textsf{ClashTypeX} & $\{ x \; \tuple{s}{m} \unif u \; \tuple{t}{m} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$ ; if $x$ and $u$ have different types \\[\jot]
  \textsf{ClashLenXF} & $\{ x \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$; if $m > n$ \\[\jot]
  \textsf{OccursCheck}     & $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \bot$;  if $x \in \Var(t)$ and $x \not= t$
\end{tabular}

@@

\hfill
\begin{tabular}{ll}
  \namedinference{SN}
  {s \eq t \mathrel\vee C \qquad u[s'] \noteq v \mathrel\vee D}
  {\sigma(u[t] \mathbin{\noteq} v \mathrel\vee C \mathrel\vee D)}
  &
  \namedinference{ER}{s \noteq s' \mathrel\vee C}{\sigma(C)} \\[2\jot]
  \namedinference{SP}
    {s \eq t \mathrel\vee C \qquad u[s'] \eq v \mathrel\vee D}
    {\sigma(u[t] \mathbin{\eq} v \mathrel\vee C \mathrel\vee D)}
  &
  \namedinference{EF}{s \eq t \mathrel\vee  s' \eq u \mathrel\vee C}{\sigma(t \noteq u \mathrel\vee  s \eq u \mathrel\vee C)}
\end{tabular}
\hfill
\kern\belowdisplayskip

@@

@@

@@

>>>>

\looseness=-1
A side effect of the flattened representation is that prefix subterms are not
shared. For example, the terms $\cst{f}\;\cst{a}$ and $\cst{f}\;\cst{a}\;\cst{b}$
correspond to the flattened cells $\cst{f}(\cst{a})$ and
$\cst{f}(\cst{a}, \cst{b}).$ The argument subterm $\cst{a}$ is shared, but not
the prefix $\cst{f}\;\cst{a}.$ Similarly,
$x$ and $x\;\cst{b}$ are represented by two distinct cells, $x()$ and
$x(\cst{b})$, and there is no connection between the two occurrences of~$x$.
%
In particular,
despite perfect sharing, their \verb|binding| fields are unconnected, leading
to inconsistencies.

@@

\vskip2\jot
\noindent
\begin{tabular}{ll}
  \textsf{Delete} & $\{ t \unif t \} \uplus \mathit{S} \Unifarrow \mathit{S}$ \\[\jot]
  \textsf{Decompose} & $\{ \cst{f} \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{m} \} \uplus \mathit{S} \Unifarrow {} 
                        \mathit{S} \cup \{ s_1 \unif t_1, \ldots, s_m \unif t_m \}$ \\[\jot]
  \textsf{DecomposeX} & $\{ x \; \tuple{s}{{m}} \unif u \; \tuple{t}{{m}} \} \uplus \mathit{S} \Unifarrow {} 
                         \mathit{S} \cup \{ x \unif u{,}\; s_1 \unif t_1, \ldots, s_{m} \unif t_{m} \}$  \\
                      & if $x$ and $u$ have the same type and $m > 0$ \\[\jot]
  \textsf{Orient}     & $\{ \cst{f} \; \overline{s} \unif x \; \overline{t} \} \allowbreak \uplus \mathit{S}
                         \Unifarrow \mathit{S} \cup \{ x \; \overline{t} \unif \cst{f} \; \overline{s} \}$ \\[\jot]
  \textsf{OrientXY}   &  $\{ x \; \tuple{s}{m} \unif y \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S}
                         \Unifarrow \mathit{S} \cup \{ y \; \tuple{t}{n} \unif x \; \tuple{s}{m} \}$  \\
                      &   if $m > n$ \\[\jot]
  \textsf{Eliminate}  & $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \{ x \unif t \} \cup \{ x \mapsto t \}(\mathit{S})$ \\
                      &   if $x \in \Var(\mathit{S}) \setminus \Var(t)$
\end{tabular}
\vskip2\jot

@@

\vskip2\jot
\noindent
\begin{tabular}{ll}
  \textsf{Clash} &   $\{ \cst{f} \; \overline{s} \unif \cst{g} \; \overline{t} \} \uplus \mathit{S} \Unifarrow \bot$; if $\cst{f} \not= \cst{g}$ \\[\jot]
  \textsf{ClashTypeX} & $\{ x \; \tuple{s}{m} \unif u \; \tuple{t}{m} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$ ; if $x$ and $u$ have different types \\[\jot]
  \textsf{ClashLenXF} & $\{ x \; \tuple{s}{m} \unif \cst{f} \; \tuple{t}{n} \} \allowbreak \uplus \mathit{S} \Unifarrow \bot$; if $m > n$ \\[\jot]
  \textsf{OccursCheck}     & $\{ x \unif t \} \uplus \mathit{S} \Unifarrow \bot$;  if $x \in \Var(t)$ and $x \not= t$
\end{tabular}
\vskip2\jot

@@

% \vskip\abovedisplayskip
\hbox{}\hfill
\begin{tabular}{ll}
  \namedinference{SN}
  {s \eq t \mathrel\vee C \qquad u[s'] \noteq v \mathrel\vee D}
  {\sigma(u[t] \mathbin{\noteq} v \mathrel\vee C \mathrel\vee D)}
  &
  \namedinference{ER}{s \noteq s' \mathrel\vee C}{\sigma(C)} \\[2\jot]
  \namedinference{SP}
    {s \eq t \mathrel\vee C \qquad u[s'] \eq v \mathrel\vee D}
    {\sigma(u[t] \mathbin{\eq} v \mathrel\vee C \mathrel\vee D)}
  &
  \namedinference{EF}{s \eq t \mathrel\vee  s' \eq u \mathrel\vee C}{\sigma(t \noteq u \mathrel\vee  s \eq u \mathrel\vee C)}
\end{tabular}
\hfill\hbox{}
\vskip\belowdisplayskip

@@

E's heuristics are largely independent of the logic used and work
unchanged for Ehoh. Yet, in preliminary experiments, we noticed that E proved
some \lfhol{} benchmarks quickly using the applicative encoding
(Sect.~\ref{sec:ehoh:introduction}), whereas Ehoh timed out.
\begin{rep}There were enough such problems to prompt us to take a
closer look. \end{rep}%
Based on these observations, we extended the heuristics\confrep{}{ to exploit
\lfhol{}-specific features}.

\ourpara{Term Order Generation}
The superposition calculus is parameterized by a term
order\begin{rep}---typically an instance of KBO or LPO\end{rep} (Sect.~\ref{sec:pre:order}).
E can generate a \emph{symbol weight} function (for KBO) and a \emph{symbol
precedence} (for KBO and LPO) based on criteria such as the symbols'
frequencies\confrep{}{, their arities,} and whether they appear in the conjecture.

In preliminary experiments, we discovered that the presence of an explicit
application operator \appvar{} can be beneficial for some problems.
\begin{rep}
%A small example will help illustrate this behavior.
Let $\cst{a} : \iota_1$,
$\cst{b} : \iota_2$, $\cst{c} : \iota_3$, $\cst{f} : \iota_1 \to \iota_2 \to
\iota_3$, $x : \iota_2 \to \iota_3$, $y : \iota_2$, and $z : \iota_3$, and
consider the clauses
%
$\cst{f} \; \cst{a} \; y \noteq \cst{c}$ and
$x \; \cst{b} \eq z$,
%
where the first one is the negated conjecture. Their applicative encoding
is
%
$\cst{@}_{\iota_2,\iota_3}(\cst{@}_{\iota_1,\iota_2\to\iota_3}(\cst{f}, \cst{a}), y) \noteq \cst{c}$ and
$\cst{@}_{\iota_2,\iota_3}(x, \cst{b}) \eq z$,
%
where $\cst{@}_{\tau,\upsilon}$ is a type-indexed family of
symbols representing the application of a function of type $\tau
\to \upsilon$.
\end{rep}%
With the applicative encoding, generation schemes can take the symbols
$\cst{@}_{\tau,\upsilon}$ into account, thereby exploiting the type
information carried by such symbols.\begin{rep} Since $\cst{@}_{\iota_2,\iota_3}$ is a
conjecture symbol, some weight generation scheme could give it a low weight,
which would also impact the second clause. By contrast, the native \lfhol{}
clauses share no symbols; the connection between them is hidden in the types
of variables and symbols, which are ignored by the heuristics.

\end{rep}

To simulate the behavior observed on applicative problems,
we introduced four generation schemes that extend E's existing
symbol-frequency-based schemes by partitioning the symbols by type. To each
symbol, the new schemes assign a frequency equal to the sum of all
symbol frequencies for its class.\confrep{}{ Each new scheme is inspired by a similarly
named type-agnostic scheme in E, without \texttt{type} in its name:
%
\begin{sloppypar}
\begin{itemize}
\item \texttt{typefreqcount} assigns as each symbol's weight the
  number of occurrences of symbols of the same type.

\smallskip
\item \texttt{typefreqrank} sorts the frequencies calculated by the function
  \texttt{typefreqcount} in increasing order and assigns each symbol a weight
  corresponding to its rank.

\smallskip
\item \texttt{invtypefreqcount} is \texttt{typefreqcount}'s inverse. If
  \texttt{typefreqcount} would assign a weight $w$ to a symbol, it assigns $M
  - w + 1$, where $M$ is the maximum symbol weight according to
  \texttt{typefreqcount}.

\smallskip
\item \texttt{invtypefreqrank} is \texttt{typefreqrank}'s inverse.
  It sorts the frequencies in decreasing order.
\end{itemize}
\end{sloppypar}}
%
\looseness=-1
We designed four more schemes (whose names begin with \verb|comb| instead of \verb|type|) that combine E's type-agnostic
and Ehoh's type-aware approaches\confrep{.}{ using a linear equation.}

To generate symbol precedences, E can sort symbols by weight
%(breaking ties by arity or the first occurrence in the input problem, for example)
and use the symbol's position in the sorted array as the basis for precedence.
To reflect the type information introduced by the applicative encoding, we
implemented four type-aware precedence generation schemes.
%\begin{rep}, called
%\texttt{typefreq}, \texttt{invtypefreq}, \texttt{combfreq}, and
%\texttt{invcombfreq}, that sort the symbols by weight according to
%\texttt{typefreqcount}, \texttt{invtypefreqcount}, \texttt{combfreqcount},
%and \texttt{invcombfreqcount}, respectively\end{rep}.
%
Ties are broken by comparing the symbols' number of occurrences and, if
necessary, the position of their first occurrence in the input.

\ourpara{Literal Selection}

The side conditions of the superposition rules SN and SP
(Sect.~\ref{sec:pre:rules}) rely on a literal
selection function to restrict the set of \emph{inference literals},
thereby reducing the search space. Given a clause, a literal selection
function returns a (possibly empty) subset of its literals. For
completeness, any nonempty subset selected must contain at least
one negative literal. If no literal is selected, all \emph{maximal}
literals become inference literals.
%Given a clause $L_1 \lor \cdots \lor L_n$, a literal selection function
%returns either $\{L_i\}$ for some $i$ or $\emptyset$.
The most widely used function is probably
\texttt{SelectMaxLComplexAvoidPosPred}, which we abbreviate to
\texttt{SelectMLCAPP}. It selects at most one negative literal, based
on size, absence of variables, and maximality of the literal in the clause.
%It
%also avoids negative literals that share a predicate symbol with a
%positive literal in the same clause.

\begin{rep}
  Intuitively, applied variables can potentially be unified with more terms than
  terms with rigid heads. This makes them prolific in terms of possible
  inference partners, a behavior we might want to avoid. On the other hand, shorter proofs
  might be found if we prefer selecting applied variables. To cover both
  scenarios, we implemented selection functions that prefer or defer selecting
  applied variables. % and that break ties using \texttt{SelectMLCAPP}.

  Let $\mathit{max}(L)=1$ if $L$ is a maximal literal of the clause it appears
  in; otherwise, $\mathit{max}(L) = 0$. Let $\mathit{appvar(L)} = 1$ if $L$ is a
  literal where either side is an applied variable; otherwise,
  $\mathit{appvar(L)} = 0$. Based on these definitions, we devised the following
  selection functions, both of which rely on \texttt{SelectMLCAPP} to break
  ties:
  %
  \begin{itemize}
  \item \texttt{SelectMLCAPPAvoidAppVar} selects a negative literal $L$ with
    the maximal value of $(\mathit{max}(L){,}\; 1-\mathit{appvar}(L))$
    according to the lexicographic order.

  \smallskip
  \item \texttt{SelectMLCAPPPreferAppVar} selects a negative literal $L$ with the
    maximal value of $(\mathit{max}(L){,}\; \mathit{appvar}(L))$
    according to the lexicographic order.
  \end{itemize}
  %
  %The presence of $\mathit{max}(L)$ as the first criterion is motivated by
  %initial experiments.
  \end{rep}

  \ourpara{Clause Selection}

  Selection of the given clause is a critical choice point.
  %To ensure completeness every candidate clause should be selected
  %eventually\confrep{}{, unless it is identified as redundant and
  %discarded before its turn has come}. Ignoring a candidate forever
  %would compromise E's completeness guarantee.
  E heuristically assigns \emph{clause priorities} and \emph{clause weights} to
  the candidates.
  The priorities provide a crude partition, whereas the weights order the clauses
  within a partition.
  E's main loop visits, in round-robin fashion, a set of priority queues. From a given
  queue, the clause with the highest priority and the smallest weight is selected.
  \begin{rep}%
  Typically, one of the queues will use the clauses' age as priority, to ensure
  fairness.
  \end{rep}

  E provides template weight functions that allow users to fine-tune parameters
  such as weights assigned to variables or function symbols. The most widely
  used template is \texttt{ConjectureRelativeSymbolWeight}\confrep{}{, which we abbreviate
  to \texttt{CRSWeight}}. It computes term and clause weights according to
  eight parameters, notably \textit{conj\_mul}, a multiplier
  applied to the weight of conjecture symbols.
  %
  % conj_multiplier, fweight, cweight, pweight, vweight, max_term_multiplier, max_literal_multiplier, pos_multiplier
  % [, app_var_multiplier]
  %
  %\begin{description}[labelwidth=\widthof{\rm\textit{appv\_mul}\,}]
  %  \item[\rm\textit{conj\_mul}]
  %    multiplier applied to the weight of conjecture symbols;
  %  \item[\rm\textit{fweight}] weight of a nonnullary function symbol;
  %  \item[\rm\textit{cweight}] weight of a nullary function symbol;
  %  \item[\rm\textit{pweight}] weight of a predicate symbol;
  %  \item[\rm\textit{vweight}] weight of a variable;
  %  \item[\rm\textit{maxt\_mul}]
  %    multiplier applied to the weight of maximal terms in a literal;
  %  \item[\rm\textit{maxl\_mul}]
  %    multiplier applied to the weight of maximal literals in a clause;
  %  \item[\rm\textit{pos\_mul}]
  %    multiplier applied to the weight of positive literals in a clause.
  %\end{description}
  %
  This template works well for some applicatively encoded problems.
  Let $\cst{a} : \iota$, $\cst{f} : \iota \to \iota$, $x : \iota$, and $y :
  \iota \to \iota$, and consider the clauses
  $y \; x \noteq x$ and
  $\cst{f} \; \cst{a} \eq \cst{a}$,
  where the first one is the negated conjecture. Their encoding
  is
  $\cst{@}_{\iota,\iota}(y, x) \noteq x$ and
  $\cst{@}_{\iota,\iota}(\cst{f}, \cst{a}) \eq \cst{a}$.
  %
  The encoded clauses share $\cst{@}_{\iota,\iota}$, whose weight
  will be multiplied by \textit{conj\_mul}---usually a factor in the
  interval $(0, 1)$. By contrast, the native \lfhol{} clauses share no symbols,
  and the heuristic would fail to notice that $\cst{f}$ and $y$ have the same
  type, giving a higher weight to the second clause.
  To mitigate this, we coded a new type-aware template,
  \CRS\texttt{TypeWeight}, that
  %\confrep{}{behaves like \CRS\texttt{Weight} except that it }
  applies the \textit{conj\_mul} multiplier to all symbols whose type
  occurs in the conjecture. \begin{rep}For the example above, since $\iota \to
  \iota$ appears in the conjecture, it would notice the relation between the
  conjecture variable $y$ and the symbol $\cst{f}$ and multiply $\cst{f}$'s
  weight by \textit{conj\_mul}.

  Natively supporting \lfhol{} allows the prover to recognize
  applied variables. It may make sense to extend clause weight templates to either
  penalize or promote clauses with such variables. To
  support this extension, we added the following parameter to \texttt{CRSTypeWeight},
  as well as to some other E's weight
  function templates:
  \textit{appv\_mul} is a multiplier applied to terms $s = x \; \tuple{t}{n}$, where $s$
    is either side of the literal and $n>0$. %, and $x$ is a variable.
  %
  In addition, we implemented a new clause priority scheme,
  \texttt{ByAppVarNum}, that separates the clauses by the number of top-level
  applied variables occurring in the clause, favoring those containing fewer
  such variables.
  \end{rep}

\ourpara{Configurations and Modes}
A combination of parameters, including term order, literal
selection, and clause selection, is called a \emph{configuration}.
%
%It is generally acknowledged that most provable problems arising in practice
%can be proved quickly in at least one configuration; but the configuration
%space being infinite, there is no easy way to find this configuration.
%
%determines whether a proof can be found in reasonable time. It is very
%laborious and difficult to manually choose correct parameters.
%
For years, E has provided an \emph{auto} mode that analyzes the input
problem and chooses a configuration known to perform well on similar problems.
More recently, E has been extended with an \emph{autoschedule} mode that
applies a portfolio of configurations in sequence on the given problem\confrep{.}{,
restarting the prover for each configuration.
%Time slicing approaches tend to perform better in practice, even if each
%configuration is given a shorter time slice.

}
Configurations that are suitable for a wide range of problems have emerged over
time. One of them is the configuration that is most often chosen by E's \emph{auto}
mode. We call it \emph{boa} (``best of \emph{auto}'')\begin{conf}.\end{conf}\begin{rep}:
%
%%% TYPESETTING: TEMPORARILY DISABLED LEFTMARGIN (?)
\begin{description}[labelwidth=\widthof{Precedence generation:}]
  \item[\rm Term order:] KBO
  \item[\rm Weight generation:] \texttt{invfreqrank}
  \item[\rm Precedence generation:] \texttt{invfreq}
  \item[\rm Literal selection:] \texttt{SelectMLCAPP}
  \item[\rm Clause selection:] ~ \\
\verb|1.CRSWeight(SimulateSOS, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
\verb|4.CRSWeight(ConstPrio, 0.1, 100, 100, 100, 100, 1.5, 1.5, 1.5),| \\
\verb|1.FIFOWeight(PreferProcessed),| \\
\verb|1.CRSWeight(PreferNonGoals, 0.5, 100, 100, 100, 100, 1.5, 1.5, 1),| \\
\verb|4.Refinedweight(SimulateSOS, 3, 2, 2, 1.5, 2)|
\end{description}
%
The clause selection scheme consists of five queues, each of which is
specified by a weight function template. The prefixes $n$\texttt{.} next to
the template names indicate that the queue will be visited $n$ times in the round-robin scheme
before moving to the next one. The first argument to each
template is the clause priority scheme.

@@

E's preprocessor transforms first-order formulas into clausal
normal form, before the main loop is started. % and inference rules are applied.
As explained in Sect.~\ref{sec:pre:clauses}, E also encodes all literals
as equations.
Beyond turning the problem into a
conjunction of disjunctive clauses, the preprocessor eliminates quantifiers,
introducing Skolem symbols for essentially existential quantifiers.

For first-order logic, skolemization preserves both satisfiability
(unprovability) and unsatisfiability (provability). In contrast, for
higher-order logics without the axiom of choice, naive skolemization is
unsound, because it introduces symbols that can be used to instantiate
higher-order variables.
%
One solution proposed by Miller \cite[Sect.~6]{dm-87-compact-proofs} is to ensure that
Skolem symbols are always applied to a minimum number of arguments.
%Accordingly, the version of \lfhol{} introduced by Bentkamp et al.\
%\cite{bbcw-21-lfho} distinguishes between mandatory and optional
%arguments.
However, to keep the implementation simple, we have decided to
ignore this issue and consider all arguments as optional, including those to
Skolem symbols. In Chapter \ref{ch:ehoh2} we extend Ehoh's logic to full higher-order logic
with the axiom of choice, which addresses the issue.
%TODO_PHD: Once Ehoh 2 paper is done, add reference to the section about choice here.

There is another transformation performed by preprocessing that is
problematic, but for a different reason. \emph{Definition unfolding} is the
process of replacing equationally defined symbols with their definitions and
removing the defining equations. A \relax{definition} is a clause of the form
$\cst{f} \; \tuple{x}{m} \eq t$, where the variables $\tuple{x}{m}$ are distinct,
$\cst{f}$ does not occur in the right-hand side $t$, and $\Var(t) \subseteq
\{x_1,\dots,x_m\}$.
This transformation preserves unsatisfiability (provability) for first-~and
higher-order logic, but not for \lfhol{}, making Ehoh incomplete. The reason
is that by removing the definitional clause, we also remove a symbol $\cst{f}$
that otherwise could be used to instantiate a higher-order quantifier.
For example, the clause set
$\{\cst{f}\;x \eq x{,}\; \cst{f}\;(y\;\cst{a}) \noteq \cst{a}\}$ is unsatisfiable,
whereas $\{y\;\cst{a} \noteq \cst{a}\}$ is satisfiable in \lfhol{}. (In full
higher-order logic, the second clause set would be unsatisfiable thanks to
the $\{y \mapsto \lambda x.\; x\}$ instance and $\beta$-conversion.)
%
For the moment, we have simply disabled definition unfolding in Ehoh. We will
enable it again once we have added support for $\lambda$-terms.

Higher-order logic treats formulas as terms of Boolean type, erasing
the distinction between terms and formulas. As a consequence, formulas might appear
as arguments not only to logical connectives but also to function symbols
or applied variables---e.g., $\cst{p} \> (\cst{a} \iand \cst{b})$,
$y \> (\inot\, \cst{a})$.
We call such formulas \emph{nested}. Kotelnikov et al.\
\cite{kotelnikov-16-fool} describe a modification to
Vampire's clausification algorithm to support nested formulas. We adapt
their approach to the clausification algorithm
\cite{nw-01-small-cnf} used by E. Given a formula $\varphi$ to
clausify, the following procedure removes nested formulas:
%
\begin{enumerate}
  \item\label{item:rec}
    Let $\chi = \varphi|_{p}$ be the leftmost outermost nested formula that is different from
    $\itrue$, $\ifalse$, or a variable $x$, if one exists;
    otherwise, skip to step~\ref{item:clausif}.
    Let $p = q.r$ where $q$ is the longest strict prefix of $p$ such that $\psi = \varphi|_q$ is a formula.
    Let $\psi' = (\chi \mathrel{\iimplies} \psi[\itrue]_r)
    \mathrel{\iand} (\inot\chi \mathrel{\iimplies} \psi[\ifalse]_r)$.
    Replace $\varphi$ by $\varphi[\psi']_q$ and repeat this step.
  \item\label{item:clausif}
    Apply all the steps of E's clausification algorithm up to and
    including skolemization.
  \item Skolemization might replace Boolean variables by new terms with
    predicate symbol heads. To remove them, follow step \ref{item:rec}.
  \item Perform the remaining steps of E's clausification algorithm,
    resulting in a set of clauses.
  \item Let $C$ be a clause that contains a literal $L$ of the form $x \eq \itrue$ or $x \noteq \itrue$, where $x$
    is a Boolean variable, if one exists; otherwise, terminate.
    Delete $C$ if it also contains the complement of $L$.
    Otherwise, replace $C$ with the clause $C[x \mapsto \ifalse]$
    if $L$ is of the form $x \eq \itrue$ and else $C[x \mapsto \itrue]$.
    Trivial literals $\ifalse \eq \itrue$ and $\itrue \noteq \itrue$ are removed from
    the resulting clause. Repeat this step.
\end{enumerate}

%Like the approach by Kotelnikov et al., our approach effectively interprets
%all by hoisting them to the level above the atom in
%which they appear. This then reduces the problem to one that first-order
%provers can understand.

As an example, consider the formula $\cst{f} \; x \mathrel{\ieq} x \mathrel{\iimplies} \cst{p} \;
(\cst{a} \iand \cst{b})$. Step~\ref{item:rec} moves
the subterm $\cst{a} \iand \cst{b}$ outward, yielding
  $\cst{f} \; x \mathrel{\ieq} x \mathrel{\iimplies}
    ((\cst{a} \iand \cst{b}) \mathrel{\iimplies} \cst{p} \>\itrue)
    \mathrel{\iand}
    ({\inot}\,(\cst{a} \iand \cst{b}) \mathrel{\iimplies} \cst{p} \>\ifalse)$.
This formula can be clausified further as usual.

\begin{sloppypar}
\begin{theorem}[Total Correctness]\,%
The above procedure always terminates and produces a set of clauses that is
equisatisfiable with the original formula $\varphi$ in \lfhol{} with
interpreted Booleans and that contains no nested formulas other than
$\itrue$, $\ifalse$, and variables.
\end{theorem}
\end{sloppypar}

\newcommand{\tmeasure}{\ensuremath{\mathcalx{W}}}

\begin{proof}
It is easy to see that steps 1,~3, and~5
produce equivalent formulas or clauses. Moreover, steps 1~and~3 remove all
offending nested formulas (i.e., other than $\itrue$,
$\ifalse$, and variables). In conjunction with the standard
clausification algorithm, which preserves and reflects satisfiability, our
procedure gives correct results when it terminates.

To prove termination, we will use a measure function~$\tmeasure$ to natural
numbers that decreases with each application of step 1~or~3.
Steps 2~and~4 rely on a terminating algorithm, whereas each
application of step~5 decreases the size of a clause. We define $\tmeasure$ by
$\tmeasure(\forall x. \> s) = \tmeasure(\exists x. \> s) = \tmeasure(s)$;
$\tmeasure(\zeta \, \tuple{s}{n}) = \sum_{i=1}^{n}{\tmeasure(s_i)}$ if
$\zeta$ is a logical connective (including $\itrue$ and
$\ifalse$); and $\tmeasure(\zeta \, \tuple{s}{n}) = 3^k (1 +
\sum_{i=1}^{n}{\tmeasure(s_i)})$ otherwise, where
$k$ is the number of offending outermost nested formulas in $\zeta \, \tuple{s}{n}$.
We must show $\tmeasure(\psi) > \tmeasure(\psi')$. By definition,
$\psi$ is of the form $\zeta \,
\tuple{s}{n}$, where $\zeta$ is not a logical connective.
Thus $\tmeasure(\psi) = 3^k (1 +
\sum_{i=1}^{n}{\tmeasure(s_i)})$. Steps 1~and~3 substitute
$\itrue$ or $\ifalse$, of measure 0, for a nested formula $\chi$
(including $\chi$'s own nested formulas) in $\psi$.
Clearly, the longer $r$ is, the more $\tmeasure(\psi')$
decreases. Taking $\left|r\right|=1$, we get the upper bound $2 \tmeasure(\chi)
+ 2 \cdot 3^{k-1} (1 + \sum_{i=1}^{n}\tmeasure(s_i) - \tmeasure(\chi))$ for
$\tmeasure(\psi')$, which is less than $\tmeasure(\psi) = 3^k (1 +
\sum_{i=1}^{n}{\tmeasure(s_i)})$.
\end{proof}

The output may contain $\itrue$, $\ifalse$, or Bool\-ean
variables as nested formulas. Since E was first developed as an untyped prover,
unification of a variable with a Boolean constant was disallowed to avoid
unsoundness. We needed to undo this in Ehoh. Ehoh must also remove trivial
literals $\ifalse \eq \itrue$ and $\itrue \noteq
\itrue$ that emerge during proof search.
\end{rep}

@@

%In this section, we consider the following questions:
How useful are Ehoh's new heuristics? And how does Ehoh perform compared with
E, used directly or in tandem with the
applicative encoding, and compared with other provers?
To answer the first question, we evaluated each new parameter independently. From
the empirical results, we derived a new configuration optimized for \lfhol{}.
For the second question, we compared Ehoh's success
rate\begin{rep} and speed\end{rep} on \lfhol{} problems with
native higher-order provers and
on applicatively encoded problems with E. We also included first-order
benchmarks to measure Ehoh's overhead.

We set a CPU time limit of 60\,s per problem.
\begin{rep}%
This is more than allotted by interactive proof tools such as Sledgehammer, or
by cooperative provers such as Leo-III and Satallax, but less than the 300\,s
of~CASC \cite{gs-17-casc}.
\end{rep}%
The experiments were performed on StarExec \cite{sst-14-starexec}
nodes equipped with Intel Xeon E5-2609\,0 CPUs clocked at 2.40\,GHz.

\ourpara{Heuristics Tuning}

We used the \emph{boa} configuration as the basis to evaluate the new
heuristic schemes. For each heuristic parameter we tuned, we changed only its
value while keeping the other parameters the same as for \emph{boa}.
\begin{rep}This gives an idea of how each parameter affects overall
performance.\end{rep}
All heuristic parameters were tested on a 5012~problem suite generated
using Sledgehammer, consisting of four variants of the Judgment Day
\cite{bn-10-sh} suite.\begin{rep} The problems were given in native
\lfhol{} syntax.\end{rep} The experiments described in this subsection were
carried out using an earlier E version (2.3).\begin{conf}
%, exploiting its support for partial application and applied variables.
Our main findings are as follows:
%
\begin{itemize}
\item The combination of the weight generation scheme \texttt{invtypefreqrank}
  and the precedence generation scheme \texttt{invtypefreq} performs best.

\smallskip
\item The literal selection heuristics \texttt{SelectMLCAPP},
  \texttt{SelectMLCAPPPreferAppVar}, and \texttt{SelectMLCAPPAvoidAppVar} give
  virtually the same results.

\smallskip
\item The clause selection function
  \texttt{ConjectureRelativeSymbolTypeWeight} with \texttt{ConstPrio} priority
  and an \textit{appv\_mul} factor of 1.41 performs best.
\end{itemize}

\looseness=-1
We derived a new configuration from \textit{boa}, called \textit{hoboa},
by enabling the features identified in the first and third points. Below, we
present a more detailed evaluation of \textit{hoboa}, along with other
configurations, on a larger benchmark suite. T%
\end{conf}%

%invtypefreqrank and invtypefreq weight and precedence generation schemes
%as the best. For clause evaluation, we picked CRSTypeWeight with ConstPrio priority and applied variable multiplier of 1.41. Lastly, for literal selection we picked SelectMLCAPPAvoidAppVar.

%\ourpara{Term Order Generation}

Evaluating the new weight and pre\-cedence generation heuristics amounted to testing
each possible combination of frequency-based schemes, including E's original
type-agnostic schemes. Figure~\ref{fig:ehoh:term-eval} shows the number of
solved (i.e., proved or disproved) problems for each combination. In this and
the following figures, the underlined number is for \emph{boa}, whereas bold
singles out the best value. In the names of the generation schemes, we abbreviated
\texttt{inv} to \texttt{i}, \texttt{type} to \texttt{t}, \texttt{freq} to
\texttt{f}, \texttt{comb} to \texttt{cm}, \texttt{count} to \texttt{cn}, and
\texttt{rank} to~\texttt{r}.

### 2022-02-23 pre/pre.tex
finished

\section{Clausal Forms}

@@

\subsection{Term Order and Selection}
Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an
ordering on terms that has the following properties:

>>>>

\section{Clausal Forms}
\label{sec:pre:clauses}

@@

\subsection{Term Order and Selection}
\label{sec:pre:order}
Superposition calculus is parametrized by a {\em reduction ordering} $\succ$, an
ordering on terms that has the following properties:
